### 2013-10-30 paper/inference.tex
adding inference section

\section{Bayesian Inference}
 \label{sec:inference}

%Inferring optimal values for the uncertain parameters is performed by solving an inverse problem 
%using observations. Specifically, the AXBT temperature data obtained during typhoon Fanapi,
%which were 
%shown to be correlated to the typhoon wind speed, are used to sharpen 
%initial estimates of the uncertain drag parameters indicated in 
%Equations (\ref{eq:prior1}-\ref{eq:prior3}).

The key idea in the Bayesian approach is to express the 
forward problem (response surface/model)
as a product of conditional probability densities. Bayes theorem 
is then used to reverse the conditioning as explained below.  
Let the observation data (AXBTs) be denoted by \{$T_i = T_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$;
their model predicted counterparts by \{$M_i= M_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$
($N$ is the number of observations) and $H$ the set of control parameters ; 
according to the Bayes' theorem we have:
\begin{equation}
 p(H| \{T_i\}_{i=1}^N) \propto 
 p(\{T_i\}_{i=1}^N | H) \ p(H)  
\label{eq:bayes}
\end{equation}
where $p(H)$ is the prior of $H$, representing the \emph{a priori} knowledge
about $H$ (before considering the observations data); 
$p(\{T_i\}_{i=1}^N | H)$ is the likelihood function representing
the probability of obtaining the data given the set of parameters $H$;
and finally $p(H| \{ T_i \}_{i=1}^N)$ is the posterior,
representing the probability that $H$ is true given the data $( \{ T_i \}_{i=1}^N )$.
In our current inference problem the set of parameters will
be taken as $H = \{ \alpha, V_{\max}, m\}$. Thus, Equation~(\ref{eq:bayes}) can be rewritten as:
\begin{equation}
p(\{ \alpha, V_{\max}, m\} | \{ T_i \}_{i=1}^N)
\propto 
p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) \ p(\alpha)p(V_{\max}) p(m)  
\end{equation}
The key point in the Bayesian inference is to define the likelihood function and 
prior \citep{sivia}. For this purpose, we define
\begin{equation} 
	 \epsilon_i = T_i - M_i \ ,
\end{equation}
where $\epsilon_i$ represents the discrepancy between the model and observations. 
We assume
that the observations are independent
and that $\epsilon_i$ is normally distributed with mean
zero and variance $\sigma^2$, i.e. 
$\epsilon_i \sim N(0,\sigma^2)$. In this 
case the likelihood function can be written as:
\begin{equation} 
p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) 
= 
\prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right)  	
\label{eq:likelihood}
\end{equation}

The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
While in general $\sigma^2$ depends on the observations, in cases where the error
amplitude is generally small and does not change throughout space and time, one may use
a simplified model and assume single hyper-parameter $\sigma^2$. In our problem, 
the typhoon wind forcing conditions vary significantly from one day to another
and AXBT data collected on each day are exposed to different measurement errors.
We thus consider different variance $\sigma^2$ for every set of AXBT data collected
on different days as indicated in Section~\ref{sec:results}\ref{sec:inverse}.

The next step is choosing the prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed:
\begin{equation} 
p(\{ \alpha, V_{\max}, m\}) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i < \{ \alpha, V_{\max}, m\} \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined in 
Equations~(\ref{eq:prior1}--\ref{eq:prior3}).
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}
Consequently, Bayes' theorem gives:
\begin{equation} 
p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N) 
\propto
\left[ \prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right) \right] 
\ p(\sigma^2)p(\alpha) p(V_{\max})p(m)
\end{equation}
where $p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N)$  is the joint posterior.

Inferring the drag coefficient parameters requires 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
sample the posterior distribution accurately and efficiently.
This MCMC phase requires repeated (tens of
thousands of)
HYCOM simulations initialized with different values of the
uncertain parameters; this step is prohibitively expensive. An
alternative is to construct a surrogate model that requires a much
smaller ensemble of HYCOM runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
PCEs to build the surrogate, which, in addition
also provide statistical properties,  such as the mean, variance and sensitivities,
efficiently.

PCE methods \citep{LeMaitreKnio2010} are based on expressing the dependence 
of a quantity of
interest, $R$ say, as a truncated expansion of the form:
\begin{equation}
  R(\xbold,t,\xxi) \doteq \sum_{k = 0}^P \hat{R}_k(\xbold,t) \Psi_k(\xxi),
\mbox{ with }
 \hat{R}_k \approx \frac{\left< R, \Psi_k \right>_Q}{\left< \Psi_k, \Psi_k \right>} .
\label{eq:stochseries}
\end{equation}
where $\hat{R}_k$ are unknown coefficients, $P$ is finite and depends on
the truncation strategy adopted, and the functions $\Psi_k(\xxi)$ form
an orthogonal basis of an underlying probability space, and $\left<
R, \Psi_k \right>_Q$ denotes an appropriate numerical quadrature.
The coefficients in the series are computed via a Non-Intrusive
Spectral Projection (NISP) as it does not require any modification to
the HYCOM code but only an ensemble of simulations. We have adopted an adaptive quadrature ~\citep{Constantine:2012,winokur:2012,conrad:2012}
to minimize the errors incurred in evaluating the
coefficients and to minimize the size of the HYCOM ensemble; furthermore
it has the virtue of automatically identifying the important stochastic
variables and steering the quadrature sampling along their directions.
Appendix C provides more details about PCEs and the adaptive quadrature.
Once the coefficients are known, the stochastic series in Equation~(\ref{eq:stochseries}) 
becomes an attractive and
faithfull surrogate for the model as it can be summed very efficiently
and at tiny fraction of the cost of a full HYCOM simulation. We will
use it to estimate the model counterparts $M_i$ of the AXBT temperatures.

>>>>

\subsection{Bayesian Inference}
 \label{sec:inference}
Bayesian inference is a statistical approach to inverse problems
that has recently gained much interest in different applications.
The key idea in this approach is to express all forms of uncertainty
in terms of probability. The first step in a Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and product of conditional probability densities. 
We then formulate the prior distribution over the unknown parameters of the model
using our best belief. Lastly, given some observation data, Bayes rule 
is used to obtain a posterior distribution for these unknown
\citep{sivia}. We review this approach briefly below.

Let the observation data (water surface elevation at the gauges) be denoted by \{$S_i = S_i(x_i,y_i,t_i)\}_{i=1}^m$;
their model predicted counterparts by \{$M_i= M_i(x_i,y_i,t_i)\}_{i=1}^m$
($m$ is the number of observations) and $\{N_i\}_{i=1}^3$ the set of 
three Manning Friction coefficients treated as the unknown parameters. 
Bayes' rule yields:
\begin{equation}
 p(\{N_i\}_{i=1}^3| \{S_i\}_{i=1}^m) \propto 
 p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(\{N_i\}_{i=1}^3)  
\label{eq:bayes}
\end{equation}
where $p(\{N_i\}_{i=1}^3)$ is the prior of $\{N_i\}_{i=1}^3$, representing the \emph{a priori} knowledge
about the manning coefficient; 
$p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3)$ is the likelihood function representing
the probability of obtaining the data given the set of parameters $\{N_i\}_{i=1}^3$;
and finally $p(\{N_i\}_{i=1}^3| \{ S_i \}_{i=1}^m)$ is the posterior,
representing the probability that $\{N_i\}_{i=1}^3$ is true given the data $( \{ S_i \}_{i=1}^m )$.

\begin{equation}
p(\{N_i\}_{i=1}^3 | \{ S_i \}_{i=1}^m)
\propto 
p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(N_1)p(N_2) p(N_3)  
\end{equation}

To formulate the likelihood function, we let 
	 $\epsilon_i = S_i - M_i $
represents the discrepancy between the model and observations 
and assume that the observations are independent
and normally distributed with mean
zero and variance $\sigma^2$, i.e. 
$\epsilon_i \sim N(0,\sigma^2)$. In this 
case the likelihood function can be written as:
\begin{equation} 
p(\{S_i\}_{i=1}^m |  \{N_i\}_{i=1}^3) 
= 
\prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right)  	
\label{eq:likelihood}
\end{equation}

The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
While in general $\sigma^2$ depends on the observations, in cases where the error
amplitude is generally small and does not change throughout space and time, one may use
a simplified model and assume single hyper-parameter $\sigma^2$.

The next step is choosing the prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed:
\begin{equation} 
p(N_i) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined in 
Equations~(\ref{eq:prior1}--\ref{eq:prior3}).
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}
Consequently, Bayes' theorem gives:
\begin{equation} 
p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m) 
\propto
\left[ \prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right) \right] 
\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
\end{equation}
where $p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m)$  is the joint posterior.

Inferring the drag coefficient parameters requires 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
sample the posterior distribution accurately and efficiently.
This MCMC phase requires repeated (tens of
thousands of) Geoclaw simulations initialized with different values of the
uncertain parameters; this step is prohibitively expensive. An
alternative is to construct a surrogate model that requires a much
smaller ensemble of GeoClaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
PCEs to build the surrogate, which, in addition
also provide statistical properties,  such as the mean, variance and sensitivities,
efficiently.

### 2013-10-30 paper/mcmc.tex
adding inference section

The observed data collected at different locations 
in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
MCMCs of $40,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure \ref{fig:mcmc} and appears to be well mixed.

@@

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 6000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.013; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ and $N_3$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$ and $N_3$.

@@

>>>>

The observed data collected at different locations 
in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure \ref{fig:mcmc} and appears to be well mixed.

@@

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.013; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ and $N_3$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$ and $N_3$.

@@

   0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835

### 2013-10-30 paper/tsunami.tex
fixed some links to figres
adding inference section

\subsection{Computational Model}
Figure~\ref{fig:setup}(left) shows the topography of the considered domain.
Land topography is shown using while the bathymetry is shown using a lighter colormap.

During this tsunami, different gauges were installed to observed the surface elevation
The locations of theses gauges are shown in Figure~\ref{setup} .
The gauges are used in the inverse problem to infer the manning coefficients.

\begin{figure}[h]
\centering
\begin{tabular}{clc}
\includegraphics[width=0.45\textwidth]{../figures/topo.pdf}  &
\includegraphics[width=0.45\textwidth]{../figures/coef.pdf} 
\label{setup}
\end{tabular}
\caption{Topography and gauge locations.}
\label{fig:setup}
\end{figure}

>>>>

\subsection{GeoClaw}
2D Shallow water equations

Physical and computational parameters
Figure~\ref{fig:setup}(left) shows the topography of the considered domain.
Land topography is shown using while the bathymetry is shown using a lighter colormap.

\begin{figure}[h]
\centering
\begin{tabular}{clc}
\includegraphics[width=0.45\textwidth]{../figures/topo.pdf}  &
\includegraphics[width=0.45\textwidth]{../figures/coef.pdf} 
\label{setup}
\end{tabular}
\caption{Topography and gauge locations.}
\label{fig:setup}
\end{figure}
\subsection{Parametric Uncertainty}
\subsection{Observations}
During this tsunami, different gauges were installed to observed the surface elevation
The locations of theses gauges are shown in Figure~\ref{setup} .
The gauges are used in the inverse problem to infer the manning coefficients.

\begin{figure}[h]      
\centering
\includegraphics[width=0.6\textwidth]{../figures/obs.pdf}
\caption{Observed data of water surface elevation with time at four different gauges.}
\label{fig:obs}
\end{figure}
